A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-05-09 graph_syntax.tex
another round of edits

\subsection{Local vs Global Nodes ((recommend omission))}

  In this graph-based syntax, a variable node is considered \emph{local} if it has exactly one outgoing connection edge or \emph{global} if it has more than one outgoing connection edge. Therefore, a variable node may be used to represent a global input if it has at least two outgoing connection edges. Otherwise, the input is formulated as a design variable that is local to a specific analysis block.
%  The set of local nodes, $V_\txt{l}$, is the collection of 
%  nodes with exactly one outgoing connection edge. 
%
%  \begin{equation}
%    V_{\txt{l}} = \{V_{\txt{var}} \st \txt{deg}^+(v)=1\}
%    \label{e:local_definition}
%  \end{equation}
%  In section \ref{s:specific vs fundamental} variable nodes that are part of 
%  an analysis block are inherently local and similarly model nodes 
%  are also local. Following the definition in Eq. \ref{e:local_definition}
%  we can now see why. Variable nodes connected to model nodes are only allowed to 
%  have one outgoing edge, and hence will always by definition be local variables.

%  The set of global nodes, $V_\txt{g}$, is the collection of all nodes with more than 
%  one outgoing connection edge. (())
%
%  \begin{equation}
%    V_{\txt{g}} = \{V_{\txt{var}} \st \txt{deg}^+(v)>1\}
%    \label{e:global_definition}
%  \end{equation}

%  Following our definition any variable node that had multiple 
%  outgoing edges, is global. In Fig. \ref{f:sellar_graph_full} 
%  the variable nodes for $x_1^*$, $z_1^*$, and $z_2^*$ have multiple outgoing 
%  edges and would be considered global. 
In the Sellar problem (Fig.~\ref{f:sellar graph}) the variable nodes $x_1^*$, $z_1^*$, and $z_2^*$ have multiple outgoing edges and are therefore considered global inputs. 
  The variable $x_1$ is usually refered to as 
  a local variable, which is contradicted by the given graph ((DP: can we cite someone calling this local?)). This discrepancy 
  arises from the explicit treatment of objectives and constraints as separate 
  expression blocks. This variable is often considered a local 
  variable because it directly affects only analysis block $A_1$. However, by expanding
  $F$ into an expression block with its own inputs, $F$ maintains its own local 
  copy of the $x_1$ variable. This formulation necessitates the creation of the global 
  node $x_1^*$ in the graph with two outgoing edges to link the two $x_1$ inputs nodes 
  in the respective blocks.

This classification of $x_1^*$ in the Sellar problem
  as a global variable agrees with the structure of the Collaborative 
  Optimization (CO) archtiecture \cite{braun1996thesis}. The rules for constructing 
  a problem with CO require that each discipline be given an independent local 
  copy of all global variables. All local variables are retained uniquely 
  within their respective disciplines, except when a local design variable appears 
  explicitely in the objective or global constraint. In this case, a global 
  variable is created and the analysis block is again assigned an 
  independent local copy of it. Following our definition of global and local variable 
  nodes, no such exception for local variables in global expressions is 
  necessary. Whenever a variable is included in an expression block, 
  it forces the creation of a global variable node in the graph which will 
automatically result in the correct CO structure.

  Analysis and expression blocks are also regarded local or global but the 
  criteria for determining this is slightly different than for variable nodes---both incoming and outgoing edges must be considered.
%  In this case, you consider the set of connection edges for the block
%  (e.g. from all the variable nodes within the block) as a set. Additionally, 
%, whereas for variables
%  only outgoing connection edges are relevant. 
  An analysis block is local if all incoming edges are from exactly one analysis 
  block and all outgoing edges are from exactly one other analysis block. A global 
  analysis block is defined as a block with incoming edges or outgoing edges 
  to more than one analysis block.   In the Sellar problem, the expression blocks for each constraint, 
  $G_1$ and $G_2$, are local to their respective analysis 
  blocks, whereas the expression block for $F$ has incoming edges from $A_1$ and 
  $A_2$, therefore it is considered global.

  Catagorizing any node or analysis block with incoming or outgoing edges
  from more than one source as global neglects a more 
  subtle aspect of some MDAO problems. Namley the locality of any node in a graph is 
  a relative property. For instance, a single node might have outgoing edges 
  to two separate analysis blocks, $A$ and $B$, but none to a third, $C$. This 
  node was global with respect to $A$ and $B$ separately, but local 
  to the group $A,B$. This situation produces a natural hierarchy in a graph, 
  from which one could form ever smaller groups as the problem is segmented into 
  finer localities.

  When solving a problem using a hierarchical decomposition approach, 
  there are various techniques for partitioning a problem into hierarchical 
  localities that are effective in different scenarios 
  \cite{krishnamachari1997optimal,michelena1997hypergraph,sobieszczanski1997,Perez2004,allison2009optimal}. 
  Tosserams \textit{et al.} proposed a language based-syntax for describing the 
  partitioning of problems, named $\Psi$\cite{tosserams2010specification}. 
  $\Psi$ provides an opposing perspective to this work, where they compose 
  hierarchies from the bottom up, into larger and larger groups. However, 
  $\Psi$ provides a compiler that can produce a standard problem representation 
  of the assembled problem and a number of example converters from that 
  standard representation to application specific formats.

\subsection{Coupling Between Analyses}
In MDAO, coupling is the mutual dependence the outputs of two or more analysis codes. In this graph-based syntax, coupling is described by the presence of a cycle between two or more analysis blocks. 
%A path, given by $P=(V_\txt{path},E_\txt{path})$, contains a cycle if at least two edges in the path refrence the same node,  $E = \{(x_0,x_1),(x_1,x_2),\ldots,(x_{k-1},x_k),(x_k,x_0)\}$. 
%Cycles decribe the reciprical dependence of a set of two or more analyses on each other's outputs. 
In the Sellar problem from Eqn \ref{eqn:simple_fpf} the two 
  coupling constraints, $H_1$ and $H_2$, provide a reciprical dependence between 
  $A_1$ and $A_2$. In a graph, these constraints appear as connection edges between 
  the outputs of each analysis block. The edges belonging to a cycle in the Sellar problem are highlighted in 
  Fig. \ref{f:sellar cycles}, where they form ((sentence is unfinished))

 \begin{figure}[htb!]
  \begin{center}
    \includegraphics[width=4.0in]{images/sellar_cycles}
  \end{center}
  \caption{Sellar problem graph with dashed edges indicating participation in a cycle.}
\label{f:sellar cycles}
\end{figure}

@@

\subsection{Multi-fidelity Problems}
  \label{ss:multi-fideliy problems}
  A multi-fidelity MDAO problem can be characterized by a formulation in which two or more different analyses 
  each calculating the same data. Multi-fidelity is represented in a graph by a variable node having an indegree greater than one, which means that multiple connection edges are directed into it. When the upper indegree limit of a variable node is set above one, then the node is implied to allow multiple fidelity instances of the variable. When the lower indegree limit of a variable node is set above one, then multiple fidelity instances are required..
%  variables, $V_{mf}$, is given as: 
%
%  \begin{equation}
%    V_{\txt{mf}} = \{ V_{\txt{var}} \st 1 < \txt{deg}^-(v) \leq \txt{deg}^-_u(v) \}
%  \end{equation}

>>>>

\subsection{Local vs Global Nodes }

  All global variable nodes have more than one outgoing edge. Conversely all local 
  variable nodes have only a single outgoing edge. Fig.~\ref{f:sellar graph} contains 
  three global variables that are not present in the problem formulation given by 
  Eq.~\ref{eqn:simple_fpf}: $x_1^*$, $z_1^*$, and $z_2^*$. The presence of the 
  new nodes is required in order to represent the equality of the given variables that
  was implied by their shared names in Eq.~\ref{eqn:simple_fpf}.

  The variable $x_1$ is usually refered to as a local variable, which is 
  contradicted by the given graph \cite{sellar1996}. This discrepancy 
  arises from the explicit treatment of objectives and constraints as separate 
  expression blocks. This variable is could be considered a local 
  variable because it directly affects only analysis block $A_1$. However, by expanding
  $F$ into an expression block with its own inputs, $F$ maintains its own local 
  copy of the $x_1$ variable. This formulation necessitates the creation of the global 
  node $x_1^*$ in the graph with two outgoing edges to link the two $x_1$ inputs nodes 
  in the respective blocks.

  This classification of $x_1^*$ in the Sellar problem
  as a global variable works well within the structure of the Collaborative 
  Optimization (CO) archtiecture \cite{braun1996thesis}. The rules for constructing 
  a problem with CO require that each discipline be given an independent local 
  copy of all global variables. All local variables are retained uniquely 
  within their respective disciplines, except when a local design variable appears 
  explicitely in the objective or global constraint. In this case, a global 
  variable is created and the analysis block is again assigned an 
  independent local copy of it. Following our definition of global and local variable 
  nodes, no such exception for local variables in global expressions is 
  necessary. Whenever a variable is included in an expression block, 
  it forces the creation of a global variable node in the graph which will 
  automatically result in the correct CO structure.

\subsection{Coupling Between Analyses}
  In MDAO, coupling is the mutual dependence the outputs of two or more analysis 
  codes. In this graph-based syntax, coupling is described by the presence of a 
  cycle between two or more analysis blocks. In the Sellar problem from Eqn \ref{eqn:simple_fpf} 
  a reciprical dependence between $D_1$ and $D_2$ exists through the variabled $y_1$ and $y_2$. 
  The edges belonging to a cycle in the Sellar problem are highlighted in 
  Fig. \ref{f:sellar cycles} where they form a closed path between analysis blocks $D_1$ and $D_2$.

  \begin{figure}[htb!]
    \begin{center}
      \includegraphics[width=4.0in]{images/sellar_cycles}
    \end{center}
    \caption{Sellar problem graph with dashed edges indicating participation in a cycle.
    \label{f:sellar cycles}
    }
  \end{figure}

@@

\subsection{Multi-fidelity Problems}
  \label{ss:multi-fideliy problems}
  A multi-fidelity MDAO problem can be characterized by a formulation in which 
  two or more different analyses each calculating the same data. Multi-fidelity 
  is represented in a graph by a variable node having an indegree greater than 
  one, which means that multiple connection edges are directed into it. When the 
  upper indegree limit of a variable node is set above one, then the node is 
  implied to allow multiple fidelity instances of the variable. When the lower 
  indegree limit of a variable node is set above one, then multiple fidelity instances are required.

### 2013-05-09 intro.tex
another round of edits

The approach begins by 
    constructing the \emph{maximal connectivity graph (MCG)} describing all possible
    interconnections between the analysis tools proposed by the engineers. Graph operations are then
    conducted to reduce the MCG down to a \emph{fundamental problem graph (FPG)} that describes the set of analysis
    tools needed to solve the specified system-level design problem.
    any particular solution procedure for the problem; any relevant MDO solution architecture, e.g. MDF,
    IDF, CO, BLISS, could be selected \emph{post facto} to implement the system-level optimization by
    constructing a \emph{problem solution graph (PSG)} and the corresponding problem solution approach.
    The concept of the FPG and the identification of feasible FPGs are the main contributions of the paper.

@@

\section{Motivation}
\subsection{Specific vs Fundamental Problem Formulation }
	\label{s:specific vs fundamental}
    The Fundamental Problem Formulation (FPF) is our terminology for a statement of the overall system-level optimization problem that does not depend on the choice of solution-specifc elements such as the MDAO framework, optimization algorithm, iterative solver, or execution sequence
    used to solve the problem. The FPF contrasts with problem statements that are written without 
    reference to a specific solution strategy, but that still imply a specific execution sequence.
    For example, consider the Sellar test problem \cite{AIAA:sellar}: 
    \begin{align}
        \txt{given} & \ \ y_1 = D_1(x_1,\hat{y_2},z_1,z_2) \notag
        \\      & \ \ y_2 = D_2(y_1,z_1,z_2) \notag
        \\\txt{min.} &\ \ F(x_1,y_1,y_2,z_2) \notag
        \\\txt{w.r.t.} & \ \ x_1,y_1,\hat{y_2},z_1,z_2 \notag
        \\\txt{s.t.} & \ \ H(y_2,\hat{y_2}) = 0 \notag 
        \\     & \ \ G_1(y_1) \geq 0 \notag
        \\     & \ \ G_2(y_2) \geq 0
        \label{eqn:simple}
    \end{align}

    Here, $D_1$ and $D_2$ represent analysis tools, $F$, $G_1$, and $G_2$ are the objective 
    and constraint functions respectively, and $H$ is the coupling constraint. 
    Equation \ref{eqn:simple} implies the execution 
    order for the problem. Analysis $D_1$ outputs $y_1$, which is an input to $D_2$. 
    Hence, $D_1$ should be run before $D_2$ with with $y_2$ iterating
    to convergence with $\hat{y_2}$. A slightly different formulation 
    is equally valid and still represents the same problem:

    \begin{align}
        \txt{given} & \ \ y_2 = D_2(\hat{y_1},z_1,z_2) \notag
        \\      & \ \ y_1 = D_1(x_1,y_2,z_1,z_2) \notag
        \\\txt{min.} &\ \ F(x_1,y_1,y_2,z_2) \notag
        \\\txt{w.r.t.} & \ \ x_1,\hat{y_1},y_2,z_1,z_2 \notag
        \\\txt{s.t.} & \ \ H(y_1,\hat{y_1}) = 0 \notag 
        \\     & \ \ G_1(y_1) \geq 0 \notag
        \\     & \ \ G_2(y_2) \geq 0
        \label{eqn:simple2}
    \end{align}

    Equation \ref{eqn:simple2} differs only slightly from Eq. \ref{eqn:simple}. 
    Notice that the formulation from Eq. \ref{eqn:simple} includes 
    $\hat{y_2}$, while Eq. \ref{eqn:simple2} includes $\hat{y_1}$. The result is that
    $D_1$ is now dependent on the output of $D_2$, and $y_1$ will be iterated on to 
    convergence with $\hat{y_1}$. Now the problem has to be solved by running $D_2$ first 
    and then $D_1$.

Equations \ref{eqn:simple} and 
    \ref{eqn:simple2}  are specific alterative expressions of a more fundamental problem description 
    that we call the FPF:

    \begin{align}
        \txt{given} & \ \ y_1 = D_1(x_1,\hat{y_2},z_1,z_2) \notag
        \\      & \ \ y_2 = D_2(\hat{y_1},z_1,z_2) \notag
        \\\txt{min.} &\ \ F(x_1,y_1,y_2,z_2) \notag
        \\\txt{w.r.t.} & \ \ x_1,\hat{y_1},\hat{y_2},z_1,z_2 \notag
        \\\txt{s.t.} & \ \ H_1(y_1,\hat{y_1}) = 0 \notag
        \\     & \ \ H_2(y_2,\hat{y_2}) = 0 \notag 
        \\     & \ \ G_1(y_1) \geq 0 \notag
        \\     & \ \ G_2(y_2) \geq 0
        \label{eqn:simple_fpf}
    \end{align}

    The formulation of Eq. \ref{eqn:simple_fpf} differs from both Eq. \ref{eqn:simple} 
    and Eq. \ref{eqn:simple2} because it includes two coupling constraints. 
The result is that Eq. \ref{eqn:simple_fpf} requires the presence of both 
    $\hat{y_1}$ and $\hat{y_2}$, which means that $D_1$ and $D_2$ each now operate with 
    their own local copies of the coupling variables. This formulation fully decouples the problem so that 
    either $D_1$ or $D_2$ could be run first or both could be run simultaneously. 
\begin{comment}
    Alexandrov and Lewis, in their work with REMS, assert that fully decoupling 
    the analyses is key to achieving reconfigurability in the problem 
    formulation\cite{alexandrov2004}. The key feature they identify to allow this decoupling 
    is the creation of independent variables for each analysis. Tosserams et. al. 
    in their development of the $\psi$ language for heirarchical problem composition
    make the same requirement that all analyses get their own local copies of 
    variables in order to use them as the fundamental building blocks for 
    problems\cite{tosserams2010specification}. Martins and Hwang derive the same requirement,
    from a different perspective, in their work on computing analytic derivatives 
    for MDAO problems\cite{Martins2012}. They determined that local copies of all variables 
    allows for derivation of all relevant methods for calculating analytic derivatives 
    from the basic chain rule.
\end{comment}

@@

    A DSM by itself is insufficient to describe complete problem formulations because it captures only information about data dependency between analyses; objective and constraint information is missing from the DSM description of the problem. 
    An alternative matrix-based syntax, called a Functional Dependency Table (FDT), was proposed by Wagner and Papalambros. 
    FDT represents the relationship between functions, including objectives and constraints, and specific variables that affect 
    them\cite{Wagner1993}. Similar to DSM, FDT also describes an adjacency matrix of a graph. Unlike the DSM graph, 
    however, the FDT graph is undirected and nodes can represent analysis tools, objectives, 
    or constraints. Edges between nodes represent a dependence on the same 
    variable. Michelena and Papalambros made use of the FDT to solve a graph partitioning problem that yielded 
    more efficient optimization problem decompositions\cite{Michelena1997}. While FDT succeeds at capturing the 
    information about objectives and constraints, its lack of directed edges 
    implies that it cannot capture the coupled data dependency that a DSM captures. 
For example, the objective, $f$, from the FPF in Eq. \ref{eqn:simple_fpf} is dependent on 
    output $m$ of analysis $A$ ((BG: I don't quite get this)). The dependence of $m$ on $A$ is now captured 
    in FDT in Fig. \ref{fig:FDT_simple} alone. This missing information implies 
    that, while FDT is very useful for partitioning problems, it is also not 
    sufficient to completely describe a problem formulation. 
    \begin{figure}[htb!]
        \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|}
            \hline
                 & $x$ & $y$ & $y^t$ & $z$ & $z^t$ & m \\ \hline
            $A$  & 1  & 1    &       &     &       &   \\ \hline
            $B$  & 1  &      &       & 1   &       &   \\ \hline
            $f$  &    &      &       &     &       & 1 \\ \hline
            $g1$ &    &      &       & 1   & 1     &   \\ \hline
            $g2$ &    & 1    & 1     &     &       &   \\
            \hline
        \end{tabular}
        \caption{Functional Dependency Table (FDT) for Eq. \ref{eqn:simple_fpf} \label{fig:FDT_simple}}
        \end{center}
    \end{figure}

>>>>

The approach begins by 
    constructing the \emph{maximal connectivity graph (MCG)} describing all possible
    interconnections between the analysis tools proposed by the engineers. Graph operations are then
    conducted to reduce the MCG down to a \emph{fundamental problem graph (FPG)} that describes the set of analysis
    tools needed to solve the specified system-level design problem.
    Any particular solution procedure for the problem; any relevant MDO solution architecture, e.g. MDF,
    IDF, CO, BLISS, could be selected \emph{post facto} to implement the system-level optimization by
    constructing a \emph{problem solution graph (PSG)} and the corresponding problem solution approach.
    The concept of the FPG and the identification of feasible FPGs are the main contributions of the paper.

@@

\section{Motivation}
\subsection{Specific vs Fundamental Problem Formulation }
	\label{s:specific vs fundamental}
    The Fundamental Problem Formulation (FPF) is our terminology for a 
    statement of the overall system-level optimization problem that does not 
    depend on the choice of solution-specifc elements such as the MDAO 
    framework, optimization algorithm, iterative solver, or execution sequence
    used to solve the problem. The FPF contrasts with problem statements that are written without 
    reference to a specific solution strategy, but that still imply a specific execution sequence.
    For example, consider the Sellar test problem \cite{AIAA:sellar}, with an FPF given as follows:

    \begin{align}
        \txt{given} & \ \ y_1 = D_1(x_1,y_2,z_1,z_2) \notag
        \\      & \ \ y_2 = D_2(y_1,z_1,z_2) \notag
        \\\txt{min.} &\ \ F(x_1,y_1,y_2,z_2) \notag
        \\\txt{w.r.t.} & \ \ x_1,y_1,y_2,z_1,z_2 \notag
        \\\txt{s.t.} & \ \ G_1(y_1) \geq 0 \notag
        \\     & \ \ G_2(y_2) \geq 0
        \label{eqn:simple_fpf}
    \end{align}

    In Eq. \ref{eqn:simple_fpf}, $y_1$ is an output of $D_1$ as well as an input 
    to $D_2$. Similarly, $y_2$ is an output of $D_2$ as well as an input 
    to $D_1$. This implies that $D_1$ and $D_2$ are coupled together by means 
    of their reciprical dependence. However no information is given as to weather
    $D_1$ or $D_2$ should be run before the other. Given Eq \ref{eqn:simple_fpf}, 
    it would be equally valid to run $D_1$ first, $D_2$ first, or both simultaneously.

    Eq. \ref{eqn:simple} gives a modified problem formulation where $D_1$ is run 
    first. A new variable is introduced, $\hat{y_2}$, to break the dependence of
    $D_1$ on $D_2$. Also a new coupling constraint, $H$, to enforce the consistency
    that would otherwise be lost due to $\hat{y_2}$. Equation \ref{eqn:simple} is 
    a specific alternative expression of the more fundamental problem description 
    given in Eq. \ref{eqn:simple_fpf}.

    \begin{align}
        \txt{given} & \ \ y_1 = D_1(x_1,\hat{y_2},z_1,z_2) \notag
        \\      & \ \ y_2 = D_2(y_1,z_1,z_2) \notag
        \\\txt{min.} &\ \ F(x_1,y_1,y_2,z_2) \notag
        \\\txt{w.r.t.} & \ \ x_1,y_1,\hat{y_2},z_1,z_2 \notag
        \\\txt{s.t.} & \ \ H(y_2,\hat{y_2}) = 0 \notag 
        \\     & \ \ G_1(y_1) \geq 0 \notag
        \\     & \ \ G_2(y_2) \geq 0
        \label{eqn:simple}
    \end{align}

@@

    A DSM by itself is insufficient to describe complete problem formulations because it 
    captures only information about data dependency between analyses; objective and 
    constraint information is missing from the DSM description of the problem. 
    An alternative matrix-based syntax, called a Functional Dependency Table (FDT), was proposed by Wagner and Papalambros. 
    FDT represents the relationship between functions, including objectives and constraints, and specific variables that affect 
    them\cite{Wagner1993}. Similar to DSM, FDT also describes an adjacency matrix of a graph. Unlike the DSM graph, 
    however, the FDT graph is undirected and nodes can represent analysis tools, objectives, 
    or constraints. Edges between nodes represent a dependence on the same 
    variable. Michelena and Papalambros made use of the FDT to solve a graph partitioning problem that yielded 
    more efficient optimization problem decompositions\cite{Michelena1997}. While FDT succeeds at capturing the 
    information about objectives and constraints, its lack of directed edges 
    implies that it cannot capture the coupled data dependency that a DSM captures. 
    The FDT in Fig. \ref{fig:FDT_simple} shows that $D_1$ is dependent on $y_2$ and 
    that $D_2$ is dependent on $y_1$, but you can not infer their coupled dependence from 
    that information alone. This missing information implies 
    that, while FDT is very useful for partitioning problems, it is also not 
    sufficient to completely describe a problem formulation. 
    \begin{figure}[htb!]
        \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|}
            \hline
                   & $x_1$ & $y_1$ & $y_2$ & $z_1$ & $z_2$ \\ \hline
            $D_1$  & 1     &       & 1     & 1     & 1     \\ \hline
            $D_2$  &       & 1     &       & 1     & 1     \\ \hline
            $F$    & 1     & 1     & 1     &       & 1     \\ \hline
            $G_1$  &       & 1     &       &       &       \\ \hline
            $G_2$  &       &       & 1     &       &       \\
            \hline
        \end{tabular}
        \caption{Functional Dependency Table (FDT) for Eq. \ref{eqn:simple_fpf} \label{fig:FDT_simple}}
        \end{center}
    \end{figure}
